---
title: "Course project Writeup"
author: "Roger Rafanell"
date: "18 de agosto de 2015"
output:
  html_document:
    pandoc_args: [
      "+RTS", "-K128m",
      "-RTS"
    ]
---

Loading cleaning and pre-processing data
------------------------------------------------

Load the training dataset **pml-training.csv**:

```{r}
library(caret)
library(ggplot2)

set.seed(32343)
setwd("/home/roger/Escritorio/predmachlearn/project/")
data <- read.csv('data/pml-training.csv', header=TRUE, sep=",")

```


After loading the training dataset, the first thing was inspect the dataset looking for *NA* values. The columns containing *NAs* have been removed:

```{r}
#Remove columns with NA
data <- data[sapply(data, function(x) !any(is.na(x)))]
```
 

We have also discarded the columns that not explains the model so much, the ones with a near zero variance, as well as the ones contaning factors which are not directly related to the numerical input data collected by the sensors:

```{r}
# get indices of data.frame columns with low variance
badCols <- nearZeroVar(data)

# remove those "bad" columns from the data set
data <- data[, -badCols]
data$cvtd_timestamp <- NULL
data$user_name <- NULL
```


The filtered dataset has been splitted in a training (75%) and testing (25%) subsets:

```{r}
inTrain <- createDataPartition(y = data$classe, p=0.75, list=FALSE)
training <- data[inTrain, ]
testing <- data[-inTrain, ]
```


In some cases, there is a need to use Principal Component Analysis (PCA) to transform the data to a smaller sub–space where the new variable are uncorrelated with one another. We performed this pre-processing technique in order to capture the features containing the 80% of the variance. 

The *preProcess* method applies (in one call) Box-Cox transformation to correct for skewness, center and to scale each variable.

```{r}
# Pre-processing: Box-Cox transformation, centered, scaled and principal component signal extraction
preProc <- preProcess(training[, -57], method=c("BoxCox", "center", "scale", "pca"), thresh= 0.80)
trainPC <- predict(preProc, training[, -57])
preProc
```

Note the PCA needed **15 components** to capture the desired variance theshold.

Data Inspection
----------------------------------------------------------------------------

Some data inspection has been done in order to plot the features (using *featurePlot* method) by presenting some some different combinations of the Principal Components (PC) **[PC1-PC7]**, **[PC8-PC15]** and **[PC1, PC8-PC15]**. The following example depicts the exploration of the **[PC1-PC7]** pairs, where some clusters could be appreciated in the **[PC1-PC2]** and **[PC2, PC3-PC7]** regions.
 
 <img class=center src=figs/PC1-PC7.png height=50>
 
This could be understood as the *big picture*, and could be used to detect some interesting patterns hidden in the data.

The following operations highlight some of the most clustered regions:
 
```{r}
# Plotting some of the relevant principal component interactions
qplot(PC1, PC2, data=trainPC, col=training$classe, xlab="PC1", ylab="PC2")
```

Fitting the model
----------------------------------------------------------------------------

Since we have found some interesting patterns, is time to train a **random forest** model. This is one of the most commonly used classification algorithm as well as the **boosting** family. The model has been resampled with 10-fold cross-validation and saved as an object to speed-up the load operations.

```{r}
# Fit the model with 10-fold cross validation
## check if model exists? If not, refit:
if(file.exists("model/modelRF.rda")) {
    ## load model
    load(file = "model/modelRF.rda")
} else {
    ## (re)fit the model and save it
    train_control <- trainControl(method="cv", number=10)
    modelFit <- train(training$classe ~ ., method="rf", data=trainPC, trControl=train_control, prox=TRUE)
    save(modelFit, file = "model/modelRF.rda")
}
```

```{r}
modelFit
```

```{r}
pred <- predict(modelFit, trainPC)
confusionMatrix(pred, training$classe)
```

```{r}
plot(main="Overall model error", modelFit$finalModel, log="y")
```

Testing the model
----------------------------------------------------------------------------

The model has been tested using a dedicated test. As it could be appreciated, the model predictions are not perfect but they are so close, reaching an *out-of-sample accuracy* of **0.977**. 

```{r}
# Apply the same PCA transformation to the testing set
testPC <- predict(preProc, testing[, -57])
```

```{r}
# Testing the predictor with the testing set
pred <- predict(modelFit, testPC)
confusionMatrix(testing$classe, pred)
```

The following figure describes the number of correct predictions *vs.* the incorrect ones (outliers). The **red** colour elements represent misclassified items whether the **green** represents correctly classified elements. The model exhibits a low classification error with new data. 

```{r}
# Plotting correct and incorrect predictions
testPC$predRight <- pred==testing$classe
qplot(PC1, PC2, colour=predRight, data=testPC, main="Newdata Prediction")
```

Validation 
----------------------------------------------------------------------------------------------------

Finally, the model has been validated against the data provided in the **pml-testing.csv** file:

```{r}
# Testing with validation data (20 samples) 
validation <- read.csv('data/pml-testing.csv', header=TRUE, sep=",")

# Remove columns with NAs
validation <- validation[sapply(data, function(x) !any(is.na(x)))]

# get indices of data.frame columns with low variance
badCols <- nearZeroVar(validation)

# remove those "bad" columns from the validation set
validation <- validation[, -badCols]
validation$cvtd_timestamp <- NULL
validation$user_name <- NULL
validationPC <- predict(preProc, validation[, -57])
results <- predict(modelFit, validationPC)
```

Predicting the following results:

```{r}
results
```
